---
title: "RDI: RNA Degradation Index"
author:
- name: Zixuan Xie
  affiliation: Crown Bioscience Inc., Suzhou, Jiangsu, P.R. China
date: "`r Sys.Date()`"
package: "RDI"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
link-citations: yes
abstract: |
  RNA degradation affects RNA sequencing (RNA-seq) data integrity and complicates transcriptomic 
  analysis. This study introduces the RNA Degradation Index (RDI), a novel measure to improve RNA 
  degradation assessment, addressing limitations of existing indices like the biased Ribosomal 
  Integrity Number (RIN) and the computationally expensive Transcript Integrity Number (TIN). 
  Leveraging on the pseudoaligner kallisto, RDI assesses RNA integrity based on the mapping depth at 
  both ends of transcripts to derive a median quality score for an RNA sample. Evaluations using mouse 
  syngeneic tumor models and public RNA-seq datasets show that RDI can evaluate RNA-seq data quality in
  an average of 8 minutes, much faster than TIN's 200 minutes, while achieving high accuracy (Spearman 
  correlation with TIN = 0.83). Application of RDI to real research data revealed that about 25% of 
  RNA-seq datasets had marked RNA degradation, which impacts transcript-level studies more severely 
  than gene-level studies. RDI's robustness and efficiency make it a valuable addition to RNA-seq data 
  analytical pipelines, helping avoid biased results from degraded samples. 
vignette: |
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{RDI: RNA Degradation Index}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
options(timeout = Inf)
knitr::opts_chunk$set(
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  tidy = FALSE,
  fig.align = 'center'
)
```

# Introduction

RNA degradation affects RNA sequencing (RNA-seq) data integrity and
complicates transcriptomic analysis. This study introduces the RNA
Degradation Index (RDI), a novel measure to improve RNA degradation
assessment, addressing limitations of existing indices like the biased
Ribosomal Integrity Number (RIN) and the computationally expensive
Transcript Integrity Number (TIN). Leveraging on the pseudoaligner
kallisto, RDI assesses RNA integrity based on the mapping depth at both
ends of transcripts to derive a median quality score for an RNA sample.

RDI works by: 1. Splitting transcript sequences into 5' and 3' regions
2. Creating a kallisto index from these split sequences 3. Quantifying
RNA-seq reads against this index 4. Calculating the ratio of 3' to 5'
coverage for each transcript 5. Deriving a sample-level RDI score based
on these ratios

# Installation

## Prerequisites

Before installing RDI, you need to have the following dependencies
installed:

1.  **kallisto**: A program for quantifying abundances of transcripts
    from RNA-seq data
    -   Download from: <https://pachterlab.github.io/kallisto/>
    -   Ensure kallisto is in your system PATH
2.  **Python 3**: Required for running the helper scripts
    -   Python 3.6 or higher is recommended
    -   Required Python packages: `argparse`, `pandas`, `numpy`

## Installing RDI

You can install RDI from the source package:

```{r install_rdi}
# Install from local source
install.packages("RDI_0.0.0.9000.tar.gz", repo=NULL, type="source")

# Or install from GitHub (if available)
# devtools::install_github("zixuanxie/RDI")
```

After installation, load the package:

```{r load_rdi}
library(RDI)
```

# Quick Start Guide

## Step 1: Create a kallisto index

First, you need to create a kallisto index from a reference
transcriptome fasta file. This step splits each transcript into 5' and
3' regions and creates an index for kallisto.

```{r create_index}
# Path to reference transcriptome fasta file
fasta_file <- "/path/to/reference_transcriptome.fa"

# Output prefix for the index
output_prefix <- "/path/to/output/index"

# Create the index
create_index(fasta_file, output_prefix)
```

This will generate two files: - `output_prefix.fa`: The split
transcriptome fasta file - `output_prefix.idx`: The kallisto index file

## Step 2: Calculate RDI

Once you have the index, you can calculate RDI for your RNA-seq samples:

```{r calculate_rdi}
# Paths to paired-end RNA-seq fastq files
fq_file1 <- "/path/to/sample_R1.fq.gz"
fq_file2 <- "/path/to/sample_R2.fq.gz"

# Path to the kallisto index created in Step 1
index_path <- "/path/to/output/index.idx"

# Number of threads to use
num_threads <- 4

# Output directory for results
output_dir <- "/path/to/output/rdi_results"

# Calculate RDI
calculate_RDI(fq_file1, fq_file2, index_path, num_threads, output_dir)
```

This will generate three files in the output directory: -
`abundance.tsv`: kallisto quantification results -
`transcripts_RDI.tsv`: RDI values for each transcript - `RDI_sum.tsv`:
Summary RDI values for the sample

# Interpreting Results

## Transcript-level RDI

The `transcripts_RDI.tsv` file contains: - `transcript_id`: The ID of
the transcript - `est_counts_ratio`: The ratio of 3' to 5' coverage
(values closer to 0 indicate more degradation) - `degradation_mode`: The
direction of degradation ("from3", "from5", "fromBoth", or "both0")

## Sample-level RDI

The `RDI_sum.tsv` file contains: - `transcript_id`: Set to
"est_counts_ratio" (placeholder) - `Index(mean)`: The mean RDI across
all transcripts - `Index(median)`: The median RDI across all transcripts
(this is the main RDI score) - `Index(stdev)`: The standard deviation of
RDI values across transcripts

### RDI Score Interpretation

-   **RDI ≈ 1**: Minimal degradation (3' and 5' coverage are balanced)
-   **RDI \< 1**: RNA degradation (3' coverage is higher than 5'
    coverage)
-   **RDI \> 1**: Unusual pattern (5' coverage is higher than 3'
    coverage)

Generally, samples with RDI \< 0.8 are considered to have significant
degradation.

# Example Workflow

Here's a complete example workflow for calculating RDI for multiple
samples:

```{r example_workflow}
# Set paths
reference_fasta <- "/path/to/human_transcriptome.fa"
index_prefix <- "/path/to/output/human_index"
samples <- list(
  sample1 = c("/path/to/sample1_R1.fq.gz", "/path/to/sample1_R2.fq.gz"),
  sample2 = c("/path/to/sample2_R1.fq.gz", "/path/to/sample2_R2.fq.gz"),
  sample3 = c("/path/to/sample3_R1.fq.gz", "/path/to/sample3_R2.fq.gz")
)
output_base <- "/path/to/output/rdi_results"

# Create index (only need to do this once)
create_index(reference_fasta, index_prefix)
index_path <- paste0(index_prefix, ".idx")

# Calculate RDI for each sample
for (sample_name in names(samples)) {
  cat(paste("Processing sample:", sample_name, "\n"))
  fq1 <- samples[[sample_name]][1]
  fq2 <- samples[[sample_name]][2]
  output_dir <- file.path(output_base, sample_name)
  
  calculate_RDI(fq1, fq2, index_path, 4, output_dir)
}

# Collect and summarize results
results <- list()
for (sample_name in names(samples)) {
  rdi_file <- file.path(output_base, sample_name, "RDI_sum.tsv")
  if (file.exists(rdi_file)) {
    rdi_data <- read.table(rdi_file, header = TRUE, sep = "\t")
    results[[sample_name]] <- rdi_data$Index.median.
  }
}

# Create summary table
summary_table <- data.frame(
  sample = names(results),
  RDI = unlist(results)
)

print(summary_table)
```

# Test Code

The RDI package includes test data and test functions to verify that the
package is working correctly. Here's how to run the tests:

## Test Data

The package includes the following test files in the `extdata`
directory: - `test.fasta`: A small test reference transcriptome -
`test_1.fq.gz`: Test RNA-seq reads (forward) - `test_2.fq.gz`: Test
RNA-seq reads (reverse)

## Running Tests

You can run the tests using the `testthat` package:

``` r
# Install testthat if not already installed
# install.packages("testthat")

# Run tests from the source directory (when developing)
library(testthat)
# If running from the source directory
test_dir("tests")

# If running from the installed package (after installation with tests)
# test_dir(system.file("tests", package = "RDI"))
```

## Test Commands

You can also run the test commands manually to verify the functionality:

``` r
# Set paths to test data
test_data_dir <- system.file("extdata", package = "RDI")
fasta_file <- file.path(test_data_dir, "test.fasta")
fq_file1 <- file.path(test_data_dir, "test_1.fq.gz")
fq_file2 <- file.path(test_data_dir, "test_2.fq.gz")

# Create output directory
output_dir <- tempdir()
index_prefix <- file.path(output_dir, "test_index")

# Test create_index function
cat("Testing create_index...\n")
create_index(fasta_file, index_prefix)

# Test calculate_RDI function
cat("Testing calculate_RDI...\n")
index_path <- paste0(index_prefix, ".idx")
rdi_output_dir <- file.path(output_dir, "rdi_test")
calculate_RDI(fq_file1, fq_file2, index_path, 1, rdi_output_dir)

# Check results
cat("Checking test results...\n")
rdi_sum_file <- file.path(rdi_output_dir, "RDI_sum.tsv")
if (file.exists(rdi_sum_file)) {
  rdi_result <- read.table(rdi_sum_file, header = TRUE, sep = "\t")
  cat("RDI test result:", rdi_result$Index.median., "\n")
} else {
  cat("Test failed: RDI_sum.tsv not found\n")
}
```

# Troubleshooting

## Common Issues

1.  **kallisto not found**: Ensure kallisto is installed and in your
    system PATH
2.  **Python not found**: Ensure Python 3 is installed and in your
    system PATH
3.  **Missing Python packages**: The install script should automatically
    install required packages, but you can install them manually with
    `pip install pandas numpy`
4.  **File not found errors**: Double-check all file paths
5.  **Permission errors**: Ensure you have write access to the output
    directories

## Debugging Tips

-   Run with verbose output to see detailed progress
-   Check the output files to ensure they're being generated correctly
-   Verify that kallisto and Python are working correctly by running
    `kallisto --version` and `python --version` in your terminal

# Session Information

```{r session_info, eval=TRUE}
sessionInfo()
```
